Accelerated convergence of convective simulations using boundary value problems

### 2018-02-15 tex/bvp_paper.tex
some work on clobbering the results section (totally rewrote it, need to do fig captions / conclusion / appendices)

\section{Results}
\label{sec:results}
While Fig. \ref{fig:time_trace} shows us that the post-BVP and post-rundown solutions
exhibit similar system energies and fluxes, it is important to examine other aspects of
the evolved solution. The BVP mainly serves to adjust the thermodynamic state of the solution;
in \RB convection, this means that the evolved temperature profile is a good measure of whether
or not the BVP achieves the proper solution.

\begin{figure}[t]
\includegraphics[width=\textwidth]{./figs/temp_comparison.png}
\caption{Comparisons of the evolved thermodynamic states of a BVP solution and a long IVP rundown at
$S = 10^{4 + 2/3}$ are shown.  (a) Evolved temperature profiles, as a function of height.
(b) The percentage difference between the temperature profiles (as shown in (a)), as a function of height.
(c) Probability distributions of point-by-point measurements of $T$ throughout the two domains over the
averaging windows are shown.  The cumulative distribution functions are overplotted and the maximum
difference between them in 0.0407. This difference arises due to the imperfect alignment of the
mean domain temperatures, but it is clear that otherwise the statistics are quite similar.
\label{fig:temp_comparison} }
\end{figure}

The temperature profiles of a 2D run at $S = 10^{4 + 2/3}$ 
for both the BVP solve and the long rundown are shown in
Fig. \ref{fig:temp_comparison}a.  While they are not perfectly aligned they are within less than
0.25\% of each other throughout the depth of the atmosphere, as shown in Fig. \ref{fig:temp_comparison}b.
While the mean temperature profile is interesting, the point-by-point fluctuations in the
temperature are what lead to convective transport, and thus examining the spread of temperature
and the allowed values that it can take are important to characterizing the system.  As seen
in Fig. \ref{fig:temp_comparison}c, aside from the noticeable difference in the mode of the
temperatures, the two runs exhibit largely the same spread of temperatures.
(AUTHOR NOTE: we aren't quite sure how to do a KS test properly here, but we can at least
list the KS statistic -- currently this is in the caption.)

In addition to examining the thermodynamics of the solutions, we are especially interested in
determining if the post-BVP solution appropriately captures key features of the nonlinear dynamics.
We show point-by-point probability distributions of the horizontal velocity, vertical velocity,
and nonlinear transport in Fig. \ref{fig:pdf_comparison}.  All of these measures have a sharp
peak at zero due to the no-slip, impenetrable boundary conditions and the denser grid sampling
near the boundaries of our Chebyshev grid.  Regardless, we see that the velocities of the 
two solutions are largely consistent.  The nonlinear transport in Fig. \ref{fig:pdf_comparison}c
shows that the rundown solution experiences a larger number of energetic, extreme events,
but the BVP displays a larger number of moderately energetic events than the IVP (as
evidenced by the differences in the cumulative distribution functions).

\begin{figure}[b]
\includegraphics[width=\textwidth]{./figs/pdf_comparison.png}
\caption{Probability distribution functions of the vertical velocity (a), horizontal velocity (b), and nonlinear
vertical transport (c) are shown for a 2D run at $S = 10^{4 + 2/3}$.  We sample flows every 0.1 time units for 500 freefall
times, and include the flows at all points on the grid at each of these samples in these distributions.  All
distributions are biased by the no-slip, impenetrable boundary conditions, coupled with the dense spacing of our
Chebyshev grids near the boundaries, so there is a large peak around zero. 
The cumulative distribution function is overplotted on each plot, with a maximum difference of
(a) 0.027, (b) 0.0061, and (c) 0.035. There is an excess of higher-flux carrying
elements in the BVP solution than in the IVP rundown, despite the rundown's distribution
having more powerful, rare events.
\label{fig:pdf_comparison} }
\end{figure}

The goodness of this method has been tested across a broad range of supercriticality.  We have
examined both Rundown and BVP solutions in the range $S = [10^{1/3}, 10^5]$ in 2D and
$S = [10^1, 10^4]$ in 3D.  We have also further explored up to $S = 10^6$
(AUTHOR NOTE: we're going up to $10^8$ in 2D right now, need to mention this is state of
the art and give refs).
We show the volume averaged Nusselt number, RMS Reynolds number, and mean temperature in
the evolved BVP solutions in in Fig. \ref{fig:parameter_space_comparison}a-c.
The relative error of the BVP measurements of these values is compared to the IVP
measurements for these values in Fig. \ref{fig:parameter_space_comparison}d-f.

In Fig. \ref{fig:parameter_space_comparison}a, we show the Nusselt number scaling in 2D and
3D.  We use a classic definition of the Nusselt number
to measure the heat transport in these systems, where
\begin{equation}
\text{Nu} = \frac{\angles{F_{\text{conv}} + F_{\text{cond}}}}{\angles{F_{\text{cond, ref}}}}
 = \frac{\angles{wT - (\text{Ra Pr})^{-1/2}\grad T}}{\angles{- (\text{Ra Pr})^{-1/2} \grad T}}.
\end{equation}
At small values of $S$, we achieve a steady state solution with a clear value of Nu.  Near
$S \sim 10^3$, our simulations begin to achieve a horizontally oscillatory state.  Due to the
use of no-slip boundaries, these states never give way to full scale shearing states, but they
do reduce the overall heat transport through the system and cause the systems to oscillate between
periods of high- and low- heat transport.  The vertical lines represent the standard deviation 
of Nu associated with these oscillations, while the error bars show the drift of the mean
value of Nu over the averaging window.  We find that $\text{Nu} \propto \text{Ra}^{1/5}$,
a weaker scaling law than that seen in traditional \RB convection \cite{johnston&doering2009},
but similar to that seen in shearing states [cite Goluskin].
(AUTHOR NOTE: Ben, I think this is entirely because of the oscillatory states.  I wonder what
would happen if we included viscous fluxes...)

In Fig. \ref{fig:parameter_space_comparison}b, we measure the rms Reynolds number, where
$\text{Re} = \angles{|\bm{u}|} / \mathcal{R}$.  We find that this scales roughly as
$\text{Re} \propto \text{Ra}^{0.45}$.

In Fig. \ref{fig:parameter_space_comparison}c, we measure the volume averaged 
temperature of our solution, then subtract out the value at the upper (fixed $T$) boundary.
If the atmosphere were perfectly isothermal, this value would be exactly zero.  Thus, this is
a measure of the temperature jump across the boundary layers, and it provides insight both
into whether or not the temperature profiles of two runs are the same and into whether
the fluxes through the two systems are the same (it is in some ways an alternate measure of Nu
[cite King 2009 supp materials]).
We find that this measure is $\propto \text{Ra}^{-1/5}$, approaching zero, as expected.

For each of these measurements, the BVP achieves the same value as the IVP to
within about 3\% (Fig. \ref{fig:parameter_space_comparison}d-f).  This is good agreement,
and the residuals have a scattered distribution of positive and negative values, indicating
that the method is not biased towards the wrong answer in a clear manner.

@@

>>>>

\section{Results}
\label{sec:results}
We study evolved solutions achieved through the SE method from convective onset
up to $S = 10^5$ in 2D and $S = 10^4$ in 3D.  These SE runs are compared to
AE runs spanning from onset up to $S = 10^7$ in 2D and $S = 10^4$ in 3D.
For a full list of simulations, we refer the reader to appendix \ref{appendix:run_table}.

We report the time- and volume-averaged values of select measurements of the
evolved solutions in Fig. \ref{fig:parameter_space_comparison}.  We report
the scaling of heat transport in the evolved solution, as quantified by the
Nusselt number, in Fig. \ref{fig:parameter_space_comparison}a.
The volume averaged Nusselt number is defined as
\begin{equation}
\text{Nu} = \frac{\angles{F_{\text{conv}} + F_{\text{cond}}}}{\angles{F_{\text{cond, ref}}}}
 = \frac{\angles{wT - \mathcal{P}\partial_z T}}{\angles{- \mathcal{P} \partial_z T}}.
\end{equation}
In 3D and in 2D when $S < 10^{3+2/3}$, the evolved system is defined by a clear
value of Nu and the convection reaches a statistically stationary state.
In 2D and at larger values of $S$, the value of Nu oscillates as a function of time
as the plumes in the solution oscillate horizontally.  Our choice of no-slip
boundary conditions prevent the fluid from entering a full-fledged shearing state
\cite{goluskin&all2014}, but the oscillatory motions which do arise cause the
system to vary between periods of low heat transport and high heat transport.
The SE simulations which span up to $S = 10^5$ exhibit the same horizontally
oscillatory motion as the AE solutions for the same initial conditions. The
scaling of the mean value of Nu is roughly $\text{Nu} \propto \text{Ra}^{1/5}$,
weaker than that reported in similar systems with fixed-T and fixed-flux boundary
conditions \cite{johnston&doering2009}.  We attribute this weaker Nu scaling to
the oscillatory nature of the plumes, which may have been avoided by previous studies
using bootstrapping techniques as initial conditions.

In Fig. \ref{fig:parameter_space_comparison}b, we report the rms Reynolds number
in the AE solutions, where
$\text{Re} = \angles{|\bm{u}|} / \mathcal{R}$.  This measure scales roughly as
$\text{Re} \propto \text{Ra}^{0.45}$, and shows little variance with time.

In Fig. \ref{fig:parameter_space_comparison}c, we report the volume averaged 
temperature of the AE solutions, with the value at the upper (fixed $T$) boundary removed.
This measurement probes the thickness of the boundary layers of the solutions, and
should show scaling which is inversely proportional to Nu in converged solutions
where fixed-flux boundary conditions are used \cite{otero&all2002}.  We find here
that $(\angles{T} - T_{\text{top}}) \propto \text{Ra}^{-1/5}$, precisely the inverse
scaling of Nu, giving confidence that these solutions are in a converged state.

In Fig. \ref{fig:parameter_space_comparison}d-f, we report the fractional difference
between measurements from the AE solutions and measurements from the SE solutions.
We find that the mean value of Nu from AE solutions is accurate to the values
from SE solutions to within $\sim 1$\%, and the same is true for $\angles{T}$ measurements.
Re measurements show slightly greater error, with AE measurements being on average
$\leq 2$\% away from the SE measurements.

@@

The measurements presented in Fig. \ref{fig:parameter_space_comparison} demonstrate
that the AE method can be powerfully employed in parameter space studies in which
large numbers of simulations are compared in a volume-averaged sense.  We now turn
our examination to a more direct comparison of AE and SE for convection at
$S = 10^5$, as has been introduced in Fig. \ref{fig:time_trace}.

As the AE method primarily serves to adjust the thermodynamic structure of the
solution, we compare the temperature profile attained by AE and SE in 
Fig. \ref{fig:temp_comparison}.  We see that the boundary layer length scale is 
nearly identical in the AE and SE solution (Fig. \ref{fig:temp_comparison}a), but that
the mean temperature in the interior differs by about 0.5\% on average
(Fig. \ref{fig:temp_comparison}b). The probability distribution function of point-by-point
temperature values are compared in Fig. \ref{fig:temp_comparison}c.  We construct
this PDF by interpolating our temperature field onto a regular grid, determing the
frequency distribution of all $T$ values, and then properly normalizing such that
the integral of the PDF is unity.  In addition to the
difference in mode which can be seen in the mean vertical profile, the SE case
has a broader spread of temperature values.  This can likely be attributed to the
slight overstabilization of the temperature profile obtained by the AE method
(as can be seen in Fig. \ref{fig:time_trace}a\&b).  One means of comparing two
probability distributions to determine if they are drawn from the same underlying
sample is through the use of a Kolmogorov-Smirnov (KS) \cite{wall&jenkins2012}.
In general, a KS test must be conducted on independent, uncorrelated data, which
poorly describes the point-by-point values of flow in a fluid simulation. Thus,
we will merely use the KS statistic, the maximum difference
between the cumulative distribution functions (CDFs) of the two sample distributions,
as a numerical method of directly comparing the two PDFs.  For the distributions
shown in Fig. \ref{fig:temp_comparison}c,
we find a KS statistic of 0.113 near the modes, which is large, and implies that
these two distributions are different.

\begin{figure}[t]
\includegraphics[width=\textwidth]{./figs/temp_comparison.png}
\caption{Comparisons of the evolved thermodynamic states of a BVP solution and a long IVP rundown at
$S = 10^{5}$ are shown.  (a) Evolved temperature profiles, as a function of height.
(b) The percentage difference between the temperature profiles (as shown in (a)), as a function of height.
(c) Probability distributions of point-by-point measurements of $T$ throughout the two domains over the
averaging windows are shown.  The cumulative distribution functions are overplotted and the maximum
difference between them in 0.0407. This difference arises due to the imperfect alignment of the
mean domain temperatures, but it is clear that otherwise the statistics are quite similar.
\label{fig:temp_comparison} }
\end{figure}

In addition to comparing the thermodynamic state achieved by the SE and AE methods,
we examine the velocities found in the evolved states.
We compute PDFs in the same manner as in Fig. \ref{fig:temp_comparison}c for the
vertical velocity (Fig. \ref{fig:pdf_comparison}a), horizontal velocity (Fig. \ref{fig:pdf_comparison}b),
and the nonlinear convective flux (Fig. \ref{fig:pdf_comparison}c). We report KS statistics
of 0.0577, 0.858, and 0.0347, respectively.  Perhaps unsurpisingly, the nonlinear
convective transport between the SE and AE methods are very similar, as is captured
in a volume-averaged sense in the Nu measurements of Fig. \ref{fig:parameter_space_comparison}a\&d).
In general, the velocity boundary conditions (Eqn. \ref{eqn:bcs}) more strongly dominate the flow
fields for the SE cases than the AE cases in Fig. \ref{fig:pdf_comparison}a\&c. 
This suggests that the velocity boundary layers are not yet fully formed in the
AE solutions. However, aside from this difference, the two solutions show a similar
distribution of realized velocity values.

\begin{figure}[b]
\includegraphics[width=\textwidth]{./figs/pdf_comparison.png}
\caption{Probability distribution functions of (a) the vertical velocity, (b) the horizontal velocity, and (c) nonlinear
convective transport are shown for a 2D runs achieved through the SE and AE methods
at $S = 10^{5}$.  Flows are sampled every 0.1 time units for 500 total time units,
and flows are interpolated onto an evenly spaced grid before sampling.
The cumulative distribution function is overplotted on each plot. 
\label{fig:pdf_comparison} }
\end{figure}

Despite differences between the SE and AE solutiosn for the case studied in 
Figs. \ref{fig:time_trace}, \ref{fig:temp_comparison}, \& \ref{fig:pdf_comparison},
the AE method is still extremely powerful.  The first application of the AE method
($t \approx 70$ in Fig. \ref{time_trace}b) immediately increases the 
average time step of our solver by a factor of 2-3. At higher supercriticality
($S = 10^7$), the AE solve immediately boosts the timestep by nearly a factor of 4.
Thus, not only does this method evolve the solution into nearly the correct state, 
but further time evolution (either to achieve precisely the correct thermodynamic
state or to take measurements of fluid quantities) happens more efficiently.
